Memory-Efficient Mixed-Precision Implementations\\for Robust Explicit Model Predictive Control

### 2019-04-11 emsoft-2019/Motiv.tex
comments for intro

% !TEX root = main.tex
\section{Motivation}\label{sec:example}

@@

As a simple example, consider the problem of designing a controller for an inverted pendulum depicted in Fig.~\ref{fig:inverted_pendulum}~{(a)}.
The goal of the controller is to keep the pendulum at the vertical position while satisfying hard constraints on the state variables and control inputs.
A model of the system can be constructed using physical principles. 
After linearization and time discretization, the model is
	\begin{equation}
		\begin{bmatrix}
			 \theta_{k+1}\\
			\omega_{k+1}
		\end{bmatrix}=
		\begin{bmatrix}
			1 & T_s\\
			\frac{T_sg}{L}& (1-\frac{T_sb}{mL^2})		
		\end{bmatrix}
		\begin{bmatrix}
			\theta_k\\
			\omega_k
		\end{bmatrix}+
		\begin{bmatrix}
			0\\
			\frac{T_s}{mL^2}
		\end{bmatrix}u_k + w_k
		\label{eq:pendul_ss}
	\end{equation}
where $\theta_k$, $\omega_k$, and $u_k$ denote respectively the angular position, angular speed, 
and the input torque at time $kT_s$ with $T_s$ being an appropriate sample time.
The disturbance $w_k\in\mathbb R^2$ bounded with $\|w_k\|\leq \Delta$ captures the modeling error due to 
linearization and discretization. 
The parameter $g=9.81 [m/s^2]$ is the gravitational acceleration, $m$ is the ball mass, $b$ is the rotational fraction coefficient, 
and $L$ is the length of the bar. 
Starting from an initial state $(\theta_0,\omega_0)$, the control goal is to converge to the equilibrium point 
$\theta=0$ and $\omega=0$.
Additionally, we require the state constraints $\theta_k\in[-\pi,\pi]$ and $\omega_k\in[-\pi/8,\pi/8]$ to hold at all time instances.

Our overall goal is to 
(i) design a robust MPC that achieves the performance objectives including hard constraints in spite of a bounded
distubrance $\Delta$ modeling the implementation error due to fixed precision;
and at the same time
(ii) to minimize the total number of bits that are required for storing the explicit controller 
while ensuring the error $e\leq \Delta$ is satisfied by the implementation.

@@

Figure \ref{fig:overview} gives a high-level overview of our proposed setup. 
We start by defining an initial bound $\Delta_0$. 
For our example, we choose $\Delta_0=0$. 
We use MATLAB to find a robust explicit MPC with the bound $\Delta_0$ on the disturbance.
Next, we ask Daisy \cite{Daisy} 
to provide us with a mixed-precision scheme such that the implementation satisfies the error bound $\Delta$, that is,
at all points in the control domain, the difference between the (infinite precision) controller function
and its implementation in the mixed-precision scheme is at most $\Delta$.

@@

Next, we ask Daisy to provide us with mixed-precision implementation for all the parameters and to compute a static bound on the
implementation error. 
The mixed-precision scheme given by Daisy results in an implementation with $6084$ bits in all 
for the corresponding robust explicit MPC.
Daisy also computes a static error bound of \RM{what?}
Since this error bound is less than our robustness margin $\Delta = 0.05$, we terminate the loop
and return this implementation.
The mixed precision implementation takes about $10\%$ less memory compared 
to the smallest uniform precision implementation that respects $\Delta$,
and about $XXX\%$ \RM{what} less memory than an implementation that uniformly uses
32 bits per constant.

>>>>

% !TEX root = main.tex
\section{Overview}\label{sec:example}

@@

As a smaller example, consider the standard problem of designing a controller for an inverted pendulum depicted in~\autoref{fig:inverted_pendulum}~{(a)}.
The goal of the controller is to keep the pendulum at the vertical position while satisfying hard constraints on the state variables and control inputs.
A model of the system can be constructed using physical principles. 
After linearization and time discretization, the model is
	\begin{equation}
		\begin{bmatrix}
			 \theta_{k+1}\\
			\omega_{k+1}
		\end{bmatrix}=
		\begin{bmatrix}
			1 & T_s\\
			\frac{T_sg}{L}& (1-\frac{T_sb}{mL^2})		
		\end{bmatrix}
		\begin{bmatrix}
			\theta_k\\
			\omega_k
		\end{bmatrix}+
		\begin{bmatrix}
			0\\
			\frac{T_s}{mL^2}
		\end{bmatrix}u_k + w_k
		\label{eq:pendul_ss}
	\end{equation}
where $\theta_k$, $\omega_k$, and $u_k$ denote respectively the angular position, angular speed, 
and the input torque at time $kT_s$ with $T_s$ being an appropriate sample time.
The disturbance $w_k\in\mathbb R^2$, bounded by $\|w_k\|\leq \Delta$, captures the modeling error due to 
linearization and discretization. 
The parameter $g=9.81 [m/s^2]$ is the gravitational acceleration, $m$ is the ball mass, $b$ is the rotational fraction coefficient, 
and $L$ is the length of the bar. 
Starting from an initial state $(\theta_0,\omega_0)$, the control goal is to converge to the equilibrium point 
$\theta=0, \omega=0$.
Additionally, we require the state constraints $\theta_k\in[-\pi,\pi]$ and $\omega_k\in[-\pi/8,\pi/8]$ to hold at all time instances.

Our overall goal is to 
(i) design a robust MPC controller that achieves the performance objectives including hard constraints in spite of a bounded
disturbance $\Delta$ modeling the implementation error due to finite precision;
and at the same time
\eva{what about the modelling error that $\Delta$ bounded in the paragraph before?}
(ii) to minimize the total number of bits that are required for storing the explicit controller 
while ensuring the error $e\leq \Delta$ is satisfied by the implementation.

@@

Figure \ref{fig:overview} gives a high-level overview of our proposed setup. 
We start by defining an initial bound $\Delta_0$. 
For our example, we choose $\Delta_0=0$. 
We use MATLAB to find a robust explicit MPC with the bound $\Delta_0$ on the disturbance.
Next, we ask Daisy
to provide us with a mixed-precision scheme such that the implementation satisfies the error bound $\Delta$, that is,
at all points in the control domain, the difference between the (infinite precision) controller function
and its implementation in the mixed-precision scheme is at most $\Delta$.

@@

Next, we ask Daisy to provide us with mixed-precision implementation for all the parameters and to compute a static bound on the
implementation error. 
The mixed-precision scheme given by Daisy results in an implementation with $6084$ bits in total 
for the corresponding robust explicit MPC.
Daisy also computes a static error bound of \RM{what?}
Since this error bound is less than our robustness margin $\Delta = 0.05$, we terminate the loop
and return this implementation.
The mixed-precision implementation takes about $10\%$ less memory compared 
to the smallest uniform precision implementation that respects $\Delta$,
and about $XXX\%$ \RM{what} less memory than an implementation that uniformly uses
32 bits.

### 2019-04-11 emsoft-2019/intro.tex
comments for intro

\section{Introduction}

Model predictive control (MPC) is a technique to design control actions by solving finite-horizon open-loop
optimal control problems at each sampling instant.
The result of each optimization gives a sequence of optimal control actions, only the first of which is applied
to the process.
The same procedure is applied in the next time instant with a shifted time horizon and a new initial state, 
after receiving the updated values of the process state.
The optimization problem in MPC uses a dynamic model of the process, encodes all input and output (state) constraints, and 
optimizes a performance index. 
MPC has shown to be successful in a wide variety of industrial applications, due to its 
ability to systematically handle processes with many state and input variables as well as constraints on them.

@@

One way to tackle these problems is through \emph{explicit MPC} \cite{Bemporad:2002,Alessio2009}, which formulates the optimization
problem but computes offline a symbolic representation of the solution as a function of the state.
At run-time, the solution is evaluated on the current state as in conventional control.
For example, for linear time invariant models with linear constraints and quadratic costs, the optimization
problem for MPC can be modeled as a quadratic program, and explicit MPC techniques solve the optimization problem
using multi-parametric programming techniques.
The explicit solution is representable as a partition of the controller domain into a number of polyhedral region,
and an affine map for each region.
The implementation of explicit MPC stores a lookup table of the affine maps and performs a number of affine computations
online to find the appropriate region for the current state and to evaluate the control.  
Explicit MPC thus expands hte class of systems being controlled by MPC strategies, by taking out the need to compute fast online
or to certify a complex optimization routine.

However, there are still bottleneks in implementing explicit MPC on resource-constrained embedded micro-controllers.
First, implementations of explicit MPCs can suffer from large memory usage, because the solution of the optimization
problem can involve many (often hundreds) of regions.
Since the memory consumption grows linearly with the number of regions, 
this can be a limited factor in using explicit MPC on resource-constrained micro-controllers.
Second, many low-end micro-controllers only support fixed point arithmetic.
Errors in the controller implementation are inversely proportional to the number of bits used for representing each variable. 
An implementation of explicit MPC has to be \emph{robust} to 
implementation errors to be able to enforce hard constraints on the states at run time.

In this paper, we consider the problem of implementing explicit MPC on low-end microcontrollers with fixed precision arithmetic
in a \emph{memory-efficient} and \emph{robust} way.
We propose an automatic controller and mixed-precision implementation co-design technique that computes a minimal set of mixed
precision assignments for all variables while ensuring the resulting implementation error remains within the robustness margin.

Our proposed method iteratively solves a robust version of the MPC which explicitly models fixed-point implementation errors
in the model; see Figure~\ref{fig:overview}.
Initially, we estimate a bound $\Delta = \Delta_0$ on the implementation error and solve a
min-max quadratic program for explicit robust MPC \cite{delaPea:2005}, where the system
model has an explicit disturbance bounded by $\Delta$.
The solution is represented as a piecewise affine map.
We use an automated mixeed precision tuning tool to find an approximate implementation of the piecewise affine
map and statically compute the error of the implementation.
If the error is at most $\Delta$, we are done: the implementation satisfies the robustness margin in the model.
If not, we increase the error bound $\Delta$ and run the loop again to find a new robust controller and find its best mixed-precision
implementation.

We have implemented our algorithm on top of Matlab's multi-parametric toolbox 
for robust explicit MPC \cite{Matlab-MPT?}
and the Daisy tool for muti-precision tuning and fixed-point error analysis \cite{Daisy}.
We have applied our technique on a number of standard benchmark examples.
Our implementation finds that our mixed-precision implementation can save up to 20\% memory (on average, XXX\%) in controller
implementations over a uniform-precision implementation (and a saving of XXX\% over a uniform 32-bit implementation), while
maintaining the correctness of the controller.
In absolute terms, the saving corresponds to YYY KB over a uniform precision implementation on our largest benchmark.
In each case, the analysis takes only a few minutes of computation.
We also demonstrate the scalability of our mixed-precision tuning and error bounds analysis:
we show that on explicit MPC controllers with thousands of regions, our tool finishes in a few minutes.

>>>>

\section{Introduction}
\eva{General comment: it feels a bit long until the intro discusses the actual problem,
but it reads nicely and motivates the problem well, so it's probably fine}
\eva{General comment: should we add some more references to related work (and not only to MPC)?}
Model predictive control (MPC) is a technique to design control actions by solving finite-horizon open-loop
optimal control problems at each sampling instant.
The result of each optimization gives a sequence of optimal control actions, only the first of which is applied
to the process.
The same procedure is applied in the next time instant with a shifted time horizon and a new initial state, 
after receiving the updated values of the process state.
The optimization problem in MPC uses a dynamic model of the process, encodes all input and output (state) constraints, and 
optimizes a performance index. 
MPC has shown to be successful in a wide variety of industrial applications, due to its 
ability to systematically handle processes with many state and input variables as well as constraints on them.

@@

One way to tackle these problems is through \emph{explicit MPC} \cite{Bemporad:2002,Alessio2009}, which formulates the optimization
problem but computes offline a symbolic representation of the solution as a function of the state.
At run-time, the solution is evaluated on the current state as in conventional control.
For example, for linear time invariant models with linear constraints and quadratic costs, the optimization
problem for MPC can be modeled as a quadratic program, and explicit MPC techniques solve the optimization problem
using multi-parametric programming techniques.
The explicit solution is representable as a partition of the controller domain into a number of polyhedral regions,
and an affine map for each region.
The implementation of explicit MPC stores a lookup table of the affine maps and performs a number of affine computations
online to find the appropriate region for the current state and to evaluate the control.  
Explicit MPC thus expands the class of systems being controlled by MPC strategies, by taking out the need to compute fast online
or to certify a complex optimization routine.

However, there are still bottleneks in implementing explicit MPC on resource-constrained embedded micro-controllers.
First, implementations of explicit MPCs can suffer from large memory usage, because the solution of the optimization
problem can involve many (often hundreds) of regions.
Since the memory consumption grows linearly with the number of regions, 
this can be a limited factor in using explicit MPC on resource-constrained micro-controllers.
Second, many low-end micro-controllers only support fixed-point arithmetic.
Errors in the controller implementation are inversely proportional to the number of bits used for representing each variable. 
An implementation of explicit MPC has to be \emph{robust} to 
implementation errors to be able to enforce hard constraints on the states at run time.

In this paper, we consider the problem of implementing explicit MPC on low-end microcontrollers with fixed-point arithmetic
in a \emph{memory-efficient} and \emph{robust} way.
We propose an automatic controller and mixed-precision implementation co-design technique that computes a minimal set\todo{What is meant by minimal?} of mixed
precision assignments for all variables while ensuring the resulting implementation error remains within the robustness margin.

Our proposed method iteratively solves a robust version of the MPC which explicitly considers fixed-point implementation errors
in the model; see~\autoref{fig:overview}.
Initially, we estimate a bound $\Delta = \Delta_0$ on the implementation error and solve a
min-max quadratic program for explicit robust MPC \cite{delaPea:2005}, where the system
model has an explicit disturbance bounded by $\Delta$.
The solution is represented as a piecewise affine map.
We use an automated mixed-precision tuning tool to find an approximate implementation of the piecewise affine
map and statically compute the error of the implementation.
If the error is at most $\Delta$, we are done: the implementation satisfies the robustness margin in the model.
If not, we increase the error bound $\Delta$ and run the loop again to find a new robust controller and find its best mixed-precision
implementation.
\eva{We bound all of the errors, including ones from choosing the wrong region. Worth mentioning here?}

We have implemented our algorithm on top of Matlab's multi-parametric toolbox 
for robust explicit MPC \cite{Matlab-MPT?}
and the Daisy tool for muti-precision tuning and fixed-point error analysis \cite{Daisy}.
We have applied our technique on a number of standard benchmark examples.
Our implementation finds that our mixed-precision implementation can save up to 20\% memory (on average, XXX\%) in controller
implementations over a uniform-precision implementation (and a saving of XXX\% over a uniform 32-bit implementation), while
maintaining the correctness of the controller.
In absolute terms, the saving corresponds to YYY KB over a uniform precision implementation on our largest benchmark.
In each case, the analysis takes only a few minutes of computation.
We also demonstrate the scalability of our mixed-precision tuning and error bounds analysis:
we show that on explicit MPC controllers with thousands of regions, our tool finishes in a few minutes\todo{Hours?}.
